---
title: "GladiaTOX: R Package for Processing High Content Screening data"
author:
- name: Vincenzo Belcastro
- name: Stephane Cano
- name: Florian Martin

package: GladiaTOX
output:
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{GladiaTOX}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
    )
```

# Introduction

`GladiaTOX` (*Belcastro et al., 2019*) is an open-source solution for HCS data processing and reporting 
that expands the `tcpl` package (toxcast pipeline, *Filer et al., 2016*). In
addition to `tcpl`'s functionalities (multiple dose-response fitting and best 
fit selection), `GladiaTOX`

- Fetches raw image quantifications via a webservice layer, also allowing 
multiple (proprietary) systems to be integrated easily
- Computes minimal effective concentrations based also on historical data
- Exports results formatted for ToxPI GUI
- Compute exposure severity scores
- Implements a suite of functionalities for quality control and processing 
reporting

## Installation and package load
```{r, eval=FALSE, include=TRUE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GladiaTOX")
BiocManager::install("stringr")
```

```{r, eval=TRUE, echo=TRUE, include=FALSE}
# Load GladiaTOX package
library(GladiaTOX)
library(stringr)
```

```{r, eval=FALSE, include=TRUE}
# Set working directory
# Here is where you need to save the raw data (see below)
working_directory = "myworkingdirectory"
setwd(working_directory)
```

<div style="position:relative; top:0px; right:0px">
```{r logo, echo=FALSE, out.width='60%'}
glPlotPieLogo()
```
</div>

## Database configuration
The `GladiaTOX` installation includes the deployment of a sqlite database 
(`sql/gladiatoxdb.sqlite` folder). This file contains the database structure 
already initialized with the necessary content needed for data processing (e.g.
processing methods entries).

The first step after database deployment is to configure access parameters. 
Let's create a copy of the database in the working directory
```{r, eval=TRUE, echo=TRUE, include=TRUE}
# Create a copy of the DB in the working directory
file.copy(file.path(system.file(package="GladiaTOX"), "sql", "gladiatoxdb.sqlite"), "./", overwrite=TRUE)
sqlite_src = "./gladiatoxdb.sqlite"
```

The `gtoxConf` configuration command below initializes all necessary variables.
```{r, eval=TRUE, echo=TRUE, include=TRUE}
# sqlite database location
gtoxConf(   drvr = "SQLite",
            host = NA,
            user = NA,
            pass = NULL,
            db = sqlite_src)
```
This database will be used in next sections to load and process a second study 
phase. The sqlite database can be seen as a sample database used for the 
following example.

## Deployed database
The deployed database already contains fully processed study, with

- `asid`: 1 (assay source id), the unique study identifier
- `asnm`: SampleStudy (assay source name), the names of the study
- `asph`: PhaseI (assay source phase), the study phase

The purpose of the call `gtoxLoadAsid()` is to list all studies available in the
database.
```{r, eval=TRUE, echo=TRUE, include=TRUE}
# List available studies
gtoxLoadAsid()
```

# Data and metadata
In this section we will explore one simple way for loading and processing HCS 
data stored in INTERVALS:

1. Access https://www.intervals.science/
2. Locate the "flavor toolbox" study
    + https://doi.org/10.26126/intervals.lwo6mb.1
3. Download the HCS dataset in your working directory
    + Datasets -> "S167360 - HCS - NHBE - RAW"
4. Unzip the content in your working directory

```{r, eval=TRUE, include=TRUE, echo=TRUE}
# Load data
mdat = read.csv(file.path("metadata", "S167360_Lung_HCS_raw_metadata.csv"))
mydat = read.csv(file.path("data", "S167360_Lung_HCS_raw_data.csv"))

# Setup study information
studyname = "FlavorToolbox"
phasename = "All"
```

## `plate`: plate metadata
The `plate` object stores metadata with plate information. 

Most of the columns have self-contained names and content; `plate` is the plate 
number (usually an integer); `tube` is the well location (H1 is row 8 column 1);
`well_type` is the content type of the well (`c` positive control, `t` 
treatment, `n` is the negative control); `endpoint` contains assay names with no
exposure duration info appended; `u_boxtrack` is a plate identifier used to join
the `plate` metadata table with the raw data table prior data is loaded in the 
GladiaTOX database.

```{r, eval=TRUE, echo=TRUE, include=TRUE}
# Build plate table
myplate = data.frame(
    "stimulus" = as.character(mdat$TestItem),
    "stimulus concentration" = paste(mdat$TestItemConcentration, "uM"),
    "exposure duration" = paste0(mdat$ExposureDuration, "h"),
    "plate" = as.numeric(as.factor(mdat$PlateID)),
    "tube" = mdat$Tube,
    "well_type" = mdat$WellType,
    "vehicle_name" = paste0("veh", mdat$VehicleID),
    "study" = studyname,
    "study.phase" = phasename,
    "cell type" = "NHBE",
    "endpoint" = mdat$EndpointCategory,
    "exposure date" = "2019-11-01",
    "plate_set" = 0,
    "Biological Replicate" = 1,
    "smkid" = "",
    "well format" = "96-well",
    "assay" = paste(mdat$EndpointCategory, paste0(mdat$ExposureDuration, "h"), sep="_"),
    "Date" = "2019-11-01",
    "u_boxtrack" = mdat$PlateID,
    check.names=FALSE, stringsAsFactors=FALSE)
myplate$stimulus[myplate$well_type=="n"] = myplate$vehicle_name[myplate$well_type=="n"]
myplate$stimulus[myplate$well_type=="c"] = "PosCTRL"
myplate["stimulus concentration"][myplate["well_type"]=="n"] = ""
```

## `chnmap`: assay metadata and channel mapping
The second metadata table contains assay mapping information.

Five endpoints are part of the cytotoxicity assay (e.g., *Cell count*, 
*membranepermeability*). Two endpoints are shown to be part of the DNA damage 
assay. Since multiple endpoints can be read from the same plate, each of them is
read on a separate channel. This column will also be used later on to join 
meatadata and data tables.

```{r, eval=TRUE, echo=TRUE, include=TRUE}
# Build channel mapping
mydat$channel = mdat$Channel[match(mydat$SampleID, mdat$SampleID)]
mychnmap = data.frame(Assay = unique(mydat[, c("EndpointCategory", "EndpointName", "channel")])$EndpointCategory,
                      Endpoint = unique(mydat[, c("EndpointCategory", "EndpointName", "channel")])$EndpointName,
                      Channel = unique(mydat[, c("EndpointCategory", "EndpointName", "channel")])$channel)
```

The content of `plate` and `chnmap` are then combined to generate the assay 
table. In the assay table, assay and endpoint are concatenated to timepoints 
to generate assays entries for the database.
```{r, echo=TRUE, eval=TRUE}
# Build assay table from plate and channel mapping file
myassay <- buildAssayTab(myplate, mychnmap)
print(head(myassay, 4), row.names = FALSE)
```
## `mydat`: image quantification raw data
The data table is an export from the image quantification instrument.

This table contains the raw fluorescence quantification values: `measure_val`; 
`rowi` and `coli` are the row and column indexes; `machine_name` is the channel 
name and is used to join this table with the assay table above; `u_boxtrack` 
is the plate identified and is used to join the table with the plate table.
```{r, eval=TRUE, echo=TRUE, include=TRUE}
print(head(mydat), row.names = FALSE)
```

# Database loading
In this sections data and metadata will be loaded in the GladiaTOX database. 

## Register study info in database
The following code will register metadata file content in the database, 
including: assays, endpoints, treatments and controls. The status of the assay 
source table (study table) before and after new study creation is displayed 
below calling `gtoxLoadAsid()`. The purpose of the call is to list all studies 
available in the database before and after the new study is added with the 
function `loadAnnot()`.
```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=TRUE}
## List of studies before loading
gtoxLoadAsid()
## Load annotation in gtoxDB
loadAnnot(myplate, myassay, NULL)
## List of studies after loading
gtoxLoadAsid()
```
The `loadAnnot` function call registers multiple study parameters in the 
database, including the creation of the new assay source id (asid). The asid 
identifies the pair study name, study phase. The asid is what will be used to 
load raw data of the study, process the study and generate reports.

## Load raw data in database
The `asid` just created can be retrieved by querying the database and specify 
the study name and phase.
```{r, echo=TRUE, eval=TRUE}
# Get assay source ID
asid = gtoxLoadAsid(fld = c("asnm", "asph"), val = list(studyname, phasename))$asid
asid
```
The `asid` and the `mydat` objects are the inputs to the `prepareDatForDB` 
function used to join metadata stored in database to the raw data stored in 
the `mydat` object.

Raw data is then loaded in the database with the `gtoxWriteData` function. 
```{r, echo=TRUE, eval=TRUE}
# Prepare and load data
# Format raw data from file and load in DB
tt = merge(mydat, mdat, by="SampleID")
mydat = data.table(
    measure_val = tt$EndpointValue,
    rowi = match(toupper(str_extract(tt$Tube, '[A-Z]+')), LETTERS),
    coli = as.numeric(str_extract(tt$Tube, '[0-9]+')),
    machine_name = tt$channel,
    u_boxtrack = tt$PlateID)
mydat <- prepareDatForDB(asid=asid, mydat)
gtoxWriteData(mydat[ , list(acid, waid, wllq, rval)], lvl = 0, type = "mc")
```
Study whose `asid` is `2` is now ready to be processed.

# Data processing and reporting
Metadata and data are now registered in the database. Next step is to select 
the processing methods we want to apply on the data. There are multiple levels 
of processing (see `gtoxLoadMthd(lvl=3)` for details). The function 
`assignDefaultMthds` is a shortcut to assign all levels methods at once. The 
methods selected would probably fit well to most users.
```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
assignDefaultMthds(asid = asid)
```
With the default selection, raw data is normalized by computing the log2 fold 
change of values in each well against the median of the corresponding controls.

## Compute the noise band
The package computes a noise band to discriminate concentration series that are 
active versus those that are not. To compute the noise band we need to process 
and normalize vehicle's data running the following code:
```{r, include=TRUE, results='hide', message=FALSE, warning=FALSE}
# Run level 1 to level 3 functions
res <- gtoxRun(asid = asid, slvl = 1, elvl = 3)
```

The default behaviour is to compute noise band margins separately for each 
endpoint. Margins correspond to 3 times the baseline median absolute deviation 
of vehicle responses. The following code computes the cutoffs and store them in 
the database.
```{r, include=TRUE, results='hide', message=FALSE, warning=FALSE}
# Extract assay endpoints ids of the study
aeidtab = gtoxLoadAeid(fld="asid", val=asid)

# Compute Vehicle Median Absolute deviation
tmp <- mapply(function(xx){
    tryCatch(gtoxCalcVmad(inputs = xx, aeid = xx, 
                          notes = "computed within study"), 
             error = function(e) NULL)},
    as.integer(aeidtab$aeid))
```

# Data processing and reporting
## Process data
Now that the user has selected the processing methods, cutoffs computed, and bad
quality plates masked, all information is in place to start processing the data
with the following command.

Once the database is populated with noise band margins, then all chemical's data
can be processed.
```{r, include=TRUE, results='hide', message=FALSE, warning=FALSE}
# Apply all functions from level 1 to level 6
res <- gtoxRun(asid = asid, slvl = 1, elvl = 6)
```

The `gtoxRun` returns a list of vectors of logical values used to check 
processing status. The resulting processed data is automatically stored in the
database along with the statistics computed. Statistics include activity
concentrations (AC10 and AC50) and minimal effective concentrations (MECs).

## Data reporting

The `gtoxReport` function, with option `type = "all"`, triggers the generation
of the full processing report. The pdf file created includes summary tables, 
dose-response curves (as seen later), and other plots for all chemicals tested
in the study (see package vignettes for details).

## Additional reporting plots
Additional reporting plots, not included in the full report, can be obtained as
described below.

MEC values can be reported in a piechart. Below an example.
```{r, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE, results='hide'}
#select endpoint and chemicals for pie chart
tf = !(grepl("NF-kB", aeidtab$aenm) |
           grepl("count", aeidtab$aenm, fixed=TRUE))
chnms = c("28-flavor mixture", "Alpha pinene", "Eucalyptol", "Matrix")
glPlotPie(asid, aeids=aeidtab$aeid[tf], chnms=chnms, 
          expos.time.ordr=c("4h", "24h"))
```

The pie plot reports the mean MECs for all endpoints measured in the study. Time
points are reported to the right (4h, 24h). Each slice is associated to an 
endpoint. Numbers on the slides indicate the corresponding MEC means.

Severity scores can also be computed and displayed. This score indicate the 
average impact of chemicals across multiple endpoints.
```{r, eval=TRUE, echo=TRUE, include=TRUE, warning=FALSE, fig.width=10, fig.height=6, fig.cap='Example of severity score plot.'}
glPlotToxInd(asid)
```
Example of severity score plot reporting the full list of chemicals used in the
study; y-axis reports the severity score value (normalized between 0 and 1); 
x-axis in just the index of the chemical.


